---
title: "Weather Station Report"
author: "Yanal Kashou - HTU"
date: "November 22, 2017"
output:
  html_document:
    css: styles.css
    df_print: default
    fig_caption: yes
    highlight: kate
    toc: yes
    toc_depth: 5
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '5'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = F,
	fig.height = 8,
	fig.width = 12,
	message = FALSE,
	warning = FALSE
)
library(xlsx)
library(knitr)
library(magrittr)
library(kableExtra)
library(corrplot)
library(ggplot2)
library(plotly)
library(grid)
library(gridExtra)
library(scales)
library(circular)
options(knitr.table.format = "html")
```

***

## The Data

Our weather station at **HTU** produces hourly weather data. The following table shows the last 3 rows of the dataframe, which are the 3 latest observations made by the weather station.

There is some data preparation to be done:  

- The *Date and Time* column must be formatted correctly.   
- The *Date* and *Time* columns are redundant and instead we will create our own *Date* and  *Time* columns by stripping our formatted *Date and Time* column.  
- Columns 12, 13 and 14 should be dropped as they simply represent system-specific units and will have no value in our analysis.  
- The *Record* column is simply an index and will also be dropped.  
- We will also rename all columns to present them legibly.  


```{r read_data}
classes <- c(Date = "Date")
df <- read.xlsx2(file = "CR1000_22_11_2017.xlsx", sheetIndex = 1, startRow = 5, colClasses = classes, stringsAsFactors = F)
df[c(2, 3, 4, 12, 13, 14)] <- NULL

colnames(df) <- c("DT", "Wind.Speed", "Wind.Direction.Degree", "Wind.Direction. std", "Solar.Rad", "Air.Temp", "Relative.Humidity", "Rain.Fall", "Visibility", "Bar.Pressure")

df$DT <- as.POSIXct(df$DT, format="%d/%m/%Y %H:%M:%S", tz="EET")
df[-1] <- data.frame(sapply(df[-1], as.numeric))
```



```{r}
df1 <- df[c(1:5)]
df2 <- df[c(6:10)]

kable(tail(df1, 3), "html") %>%
   kable_styling("striped", full_width = T)

kable(tail(df2, 3), "html") %>%
    kable_styling("striped", full_width = T)

```

***

## Discovering Correlations

### Correlations Between Features
First of all, we are going to plot a correlation heatmap between selected features. Namely, *Wind Speed*, *Wind Direction*, *Solar Radiation*, *Air Temp*, *Relative Humidity*, *Rain Fall* and *Visibility*.

This is done to discover correlations between the features that the station collected.

```{r correlation_features}
df_cor <- df[,-1]
df_cor <- df_cor[, -9]
df_cor <- df_cor[,-c(2, 3)]
# colnames(df_cor) <- c("Wind Speed", "Wind Direction", "Solar Radiation", "Air Temp.", "Relative Humidity", "Rain Fall", "Visibility")
df_cor <- data.frame(sapply(df_cor, as.numeric))
M <- cor(df_cor)
corrplot(M, method = "number")
```

There are some correlations in our data, specifically we can see that *Air Temperature* has the highest positive correlation with *Solar Radiation*, we also see that *Relative Humidity* has the highest negative correlation with *Air Temperature*, which is logical to deduce.

### Correlations with Time

In the previous section, we discovered some direct correlations between features, however, there is no indication of high correlations of any variable with *Visibility*. We would need to take a deeper exploratory look into the data, specifically in the time-domain, to see if there are indeed any hidden or indirect correlations.

Below we plot the Time and two variables, `Solar.Rad` and the logarithm of `Visibility`. We will also introduce regression lines for each. Noteably, using the logarithmic scale with a regression smoother that captures only 25% (red) of the data shows us a general trend in decrease in `Visibility`. And this corresponds with the local decrease in `Solar.Rad`at the end of the time period.

```{r corr_time_a}
a <- ggplot(df, aes(DT, Solar.Rad)) +
  geom_line() + 
  scale_x_datetime(breaks = date_breaks("1 week"), labels = date_format("%d/%m %I:%M %p")) +
  stat_smooth(span = 0.075, color = "orange") +
  xlab("Time") +
  ylab("Solar Radiation") +
  ggtitle("Plot of Solar Radiation in (kW/m^2) with Time (h) ") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme_minimal()
# p1 <- ggplotly(a)
```

```{r plot corr_time_b}
b <- ggplot(df, aes(DT, log(Visibility))) +
  geom_line() + 
  scale_x_datetime(breaks = date_breaks("1 week"), labels = date_format("%d/%m %I:%M %p")) +
  stat_smooth(span = 0.075) +
  stat_smooth(span = 0.75, colour = "red") +
  xlab("Time") +
  ylab("log(Visibility)") +
  ggtitle("Plot of Logarithmic Visibility in (M) with Time (h) ") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme_minimal()
# p2 <- ggplotly(b)

# subplot(p1, p2, nrows = 2, margin = 0.04)
grid.arrange(a, b, nrow=2)
```

***

## Sustained Wind Speeds
As above with the temporal plots, we can use the logarithm of the `Wind.Speed` plotted against `Time`, along with different regression lines with 25% (blue), 50% (purple), and 75% (black) of the data to capture a trend.

```{r sustained_wind}
c <- ggplot(df, aes(DT, log(Wind.Speed))) +
  geom_point(colour = "black") + 
  scale_x_datetime(breaks = date_breaks("1 week"), labels = date_format("%d/%m %I:%M %p")) +
  stat_smooth(span = 0.25, color = "blue") +
  stat_smooth(span = 0.50, color = "purple") +
  stat_smooth(span = 0.75, color = "black", size = 1.5) +
  xlab("Time") +
  ylab("Speed in KM/H") +
  ggtitle("Plot of Sustained Wind Speed over Time ") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme_minimal()
c
```

We can clearly see a general rise in Wind Speed over the period examined. And logarithmically speaking, it is an order of magnitude greater than the beginning of the period.

***

## General Wind Direction

```{r polar}
wind.density <- density(as.circular(df$Wind.Direction.Degree,
                                    control.circular = list(units = "degrees")), 
                        bw = 30)
wind.density.df <- as.data.frame(wind.density[c("x", "y")])

ggplot(wind.density.df, aes(x, y)) +
  geom_area() +
  scale_x_continuous(breaks = c(0, 90, 180, 180 + 90), 
                     labels = c("North", "East", "South", "West")) +
  coord_polar() +
  theme_bw() + 
  ggtitle("General Wind Direction")
```

We can deduce from this awkward looking plot that most of the wind either blew from the north or from the south-west, while some of the wind can be attributed to the north-west.


